---
title: "Flow_cytometry_analysis"
author: "Adam Sorbie"
date: "2024-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

Changes to the code are only required in this code block, and are marked
with a comment


```{r}
source("helper_functions.R")
pacman::p_load("tidyverse", "ggprism", "rstatix", "ggpubr", "EnvStats", "janitor", "readxl", "pheatmap")
conflicted::conflicts_prefer(rstatix::filter)

# LOAD REAL DATA HERE - MUST CONTAIN % COLUMNS FORMATTED ACCORDING TO "percentage_calculation_database.xlsx" FILE
flowjo_dat <- read_xlsx("Brain_S4.xlsx", sheet = 1)

# load calculation database 
calc_db <- read_xlsx("percentage_calculation_database S4.xlsx")
conflicted::conflicts_prefer(rstatix::select)
```

Force first column name to be sample ID

```{r}
flowjo_dat <- flowjo_dat %>% 
  rename(sampleID = 1)
```

Wrap these into functions when refactoring


```{r}
populations <- flowjo_dat %>% select(contains("%")) %>% colnames()
calc_db_filt <- calc_db %>% filter(population_perc %in% populations)


extra_columns <- setdiff(populations, calc_db$population_perc)
if (length(extra_columns) > 0) {
  cat("The following populations are not in the database:\n", paste(extra_columns, collapse = ", "), "\n")
}

```

Check columns necessary for calculation are named correctly 

```{r}
key_cols <- c("beads concentration", "beads | count", "Live cells | count")
if (! all(key_cols %in% colnames(flowjo_dat))){
  print(paste("Columns", key_cols, "not found in data, please rename or add these columns to dataframe"))
}
```


```{r}
# remove percentages
flowjo_dat <- flowjo_dat %>% 
  select(-contains("%"))

calc_cols <- flowjo_dat %>% 
 select(`beads concentration`:last_col(), -c(`beads concentration`, `beads | count`)) %>% 
  colnames()
```

```{r}
flowjo_dat_calc <- flowjo_dat %>% 
  mutate(across(.cols = all_of(calc_cols), .fns=~ (.x / `beads | count`) * `beads concentration`, .names = "{.col}_adj")) 
```

Need to throw an error here if column in calc_db missing 
```{r}
pairs <- list()
for (i in 1:nrow(calc_db_filt)){
  pairs[[i]] <- c(as.character(calc_db_filt[i,]["population"]), as.character(calc_db_filt[i,]["parent"]))
}
names(pairs) <- calc_db_filt$population_perc


flowjo_dat_calc_perc <- map_dfc(pairs, ~ flowjo_dat_calc %>% 
                select(contains(.x)) %>% 
                reduce(., `/`) * 100) %>%
    rename_all(~ names(pairs)) %>%
    bind_cols(flowjo_dat_calc, .)
```

Flag outliers

Think about a better way to do this - 
* could print flagged samples and population
* could bind rows and return dataframe summary and print nicely with e.g. kable 

```{r}
outlier_dat <- flowjo_dat_calc_perc %>% 
    select(sampleID, contains("_adj"))
adj_columns <- colnames(flowjo_dat_calc_perc %>% 
    select(contains("_adj")))
for (i in adj_columns){
  
  outlier_det <- identify_outliers(outlier_dat, variable = i)
  if (nrow(outlier_det) > 0){
    print(outlier_det %>% select(sampleID, i, is.outlier))
  }
}
```
### Remove outliers 

can't simply filter here, would need to just NA value 
```{r}
#flowjo_dat_calc_perc <- 
```


Calculate stats across all count_adj and % columns


### Absolute counts 

```{r}
absolute_counts <- flowjo_dat_calc_perc %>% 
  select(Condition, sampleID, contains("count_adj")) %>% 
  # need to make this flexible
  column_to_rownames("sampleID")

heatmap_annotation <- flowjo_dat %>% 
  select(sampleID, Condition) %>% 
  column_to_rownames("sampleID")
```

### Outlier flagging

Heatmap 

Ideally we do show colnames, need fix in-case they are super long
```{r, fig.height=10, fig.width=12}
pheatmap(absolute_counts %>% select(-Condition) %>% t() %>% log10(), cluster_cols = F, annotation_col = heatmap_annotation)
```

```{r, fig.height=6.5, fig.width=5.5}
# have to grab condition on the fly 
absolute_counts_long <- absolute_counts %>% 
  pivot_longer(-Condition, names_to = "marker")

comparisons <- make_combinations(absolute_counts_long$Condition)
for (i in unique(absolute_counts_long$marker)){
  
  sub <- absolute_counts_long %>% 
    filter(marker == i)
  
  marker <- sub("\\|.*", "", i)

  p <- ggbarplot(sub, x = "Condition", y="value", fill = "Condition", 
          add = c("mean_se"), ylab = paste0(marker, " (Absolute count)"),
          width=0.5) +
  geom_jitter(width = 0.1, size=3) + 
     scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  scale_color_prism("floral") +
  scale_fill_prism("floral") +
  theme_prism(base_size = 16) +
  theme(legend.position = "none") +
  stat_compare_means(comparisons = comparisons, method = "wilcox.test", label = "p.format") 
  print(p)
}
```

### Percentages

```{r}
# need to also figure out fix for sample id column
perc <- flowjo_dat_calc_perc %>% 
  select(Condition, sampleID, contains("%")) %>% 
  column_to_rownames("sampleID")
perc_long <- perc %>% 
  pivot_longer(-Condition, names_to = "marker")
```


```{r, fig.height=10, fig.width=12}
pheatmap(perc %>% select(-Condition) %>% t(), cluster_cols = F, annotation_col = heatmap_annotation)
```


```{r, fig.height=6.5, fig.width=5.5}
for (i in unique(perc_long$marker)){
  
  sub <- perc_long %>% 
    filter(marker == i)
  
  marker <- sub("\\|.*", "", i)

  p <- ggbarplot(sub, x = "Condition", y="value", fill = "Condition", 
          add = c("mean_se"), ylab = paste0(marker),
          width=0.5) +
  geom_jitter(width = 0.1, size=3) + 
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  stat_compare_means(comparisons = comparisons, method = "wilcox.test", label = "p.format") + 
  scale_color_prism("floral") + 
  scale_fill_prism("floral") + 
  theme_prism(base_size = 16) + 
  theme(legend.position = "none") 
  print(p)
}

  
```